$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-09-03 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
corrected typos

The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters. We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing. Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}). Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.

@@

In \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain convergence is diagnosed. We illustrate this via four examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; (c), a bivariate normal with a non-stationary covariance; and (d), an autocorrelated sampling distribution with a univariate normal target with a range of different autocorrelations. The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue. The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish as the numbers of dimensions considered increased. Example (c) showed that split-$R^*$ opposed to split-$\widehat{R}$ is able to diagnose non-stationary covariance between dimensions of a target distribution. In (d), we show that $R^*$ is able to differentiate between distributions with non-stationary target distributions and stationary ones. It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there insufficient draws.

>>>>

The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters. We split each chain's draws into randomly divided training and testing sets: here, we use 70\% of draws for training and 30\% for testing. Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}). Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.

@@

In \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain convergence is diagnosed. We illustrate this via four examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; (c), a bivariate normal with a non-stationary covariance; and (d), an autocorrelated sampling distribution with a univariate normal target with a range of different autocorrelations. The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue. The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish as the numbers of dimensions considered increased. Example (c) showed that split-$R^*$ opposed to split-$\widehat{R}$ is able to diagnose non-stationary covariance between dimensions of a target distribution. In (d), we show that $R^*$ is able to differentiate between distributions with non-stationary target distributions and stationary ones. It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there are insufficient draws.
